Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2021-07-08 paper-correlations.tex
Introduction

\begin{document}
%*Corresponding author: Kirill M. Gerke, Tel.: +79661877715, E-mail: kg@ifz.ru,
%Address: Schmidt Institute of Physics of the Earth of Russian Academy of
%Sciences, Bolshaya Gruzinskaya str. 10/1, Moscow, 123242, Russia
\author{Vasily Postnicov}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of Sciences, Moscow, 123242, Russia}
\author{Alexey I. Samarin}
\altaffiliation{Computational Mathematics and Cybernetics, Lomonosov Moscow State University, Moscow, 119991, Russia}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of Sciences, Moscow, 123242, Russia}
\author{Marina V. Karsanina}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of Sciences, Moscow, 123242, Russia}
\author{Efim V. Lavrukhin}
\altaffiliation{Computational Mathematics and Cybernetics, Lomonosov Moscow State University, Moscow, 119991, Russia}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of Sciences, Moscow, 123242, Russia}
\author{Kirill M. Gerke}
\email[E-mail:]{kg@ifz.ru}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of Sciences, Moscow, 123242, Russia}

\title{Something about correlation functions}

@@

@@

\section{Verifying our algorithms}
To verify our package we use random datasets of overlapping disks (the number of
dimensions $n = 2$) or balls ($n = 3$) with radii $R$ and centers placed in
points generated by Poisson point process with a parameter $\lambda$. For these
datasets correlation functions can be given in closed-form expressions for one-,
two- and three-dimensional cases. For the sake of testing we obtain a collection
of disks (resp. balls) in a square (resp. cube) with the side $s$ using the
following algorithm:
\begin{algorithmic}[1]
  \label{testdata}
  \Procedure{testdata}{$s, n, R, \lambda$}
    \State $nballs \gets poisson(\lambda s^n)$ \Comment{$poisson(\lambda)$
      generates random numbers with Poisson distribution with parameter
      $\lambda$.}
    \State Generate $nballs$ random points which are uniformly distributes in
    a cube ($n = 3$) or square ($n = 2$) with a side $s$
    \State Draw balls (disks) of radius $R$ in those points. Store the result in
    an array $output$.
    \State \textbf{return} $output$
  \EndProcedure
\end{algorithmic}

@@

\section{Efficiency of our algorithms}

To measure efficiecy of our algorithms we use a machine with the following
configuration:
\begin{itemize}
\item \textbf{CPU}: AMD Ryzen 5 1600X running at 2200~MHz
\item \textbf{Memory}: 32~Gb DDR4-2400 memory
\item \textbf{GPU}: Nvidia Tesla V100 with 32~Gb video memory
\end{itemize}

@@

@@

\begin{algorithmic}[1]
  \Procedure{countruns}{$array, len, phase$}
    \State $runslist \gets [\quad]$
    \State $runs \gets 0$
    \ForAll{$x \in array$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$array, runs, n$}
    \For{$idx = 1,n$}
      \State $array[idx] \gets array[idx] + runs$
      \State $runs \gets runs - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$array, first, last, n$}
    \State $sum \gets first + last$
    \State $n \gets \min(n, sum)$
    \For{$idx = 1,n$}
      \State $a \gets \min(idx - 1, first, last, sum - (idx - 1))$
      \State $array[idx] \gets array[idx] + a$
    \EndFor
  \EndProcedure
  \\
  \Procedure{l2}{$array, phase, maxlen, periodic$}
    \State $success \gets zeros(maxlen)$
    \State $total \gets zeros(maxlen)$
    \State $len \gets length(array)$
    \State $nupdates \gets \min(len, maxlen)$
    \State $runs \gets countruns(array, maxlen, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(success, run, \min(run, maxlen))$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(array)$
      \State $last \gets last(array)$
      \If{$first = last$}
        \State $updateperiodic(success, first, last, nupdates)$
      \EndIf
      \State Increment $total[\text{from} \, 1 \, \text{to} \, nupdates]$ by
      $len$
    \Else
      \State $updateruns(total, len, nupdates)$
    \EndIf
    \State \textbf{return} $success$ divided by $total$ element-wise.  
  \EndProcedure
\end{algorithmic}
\end{document}

>>>>

\begin{document}
%*Corresponding author: Kirill M. Gerke, Tel.: +79661877715, E-mail: kg@ifz.ru,
%Address: Schmidt Institute of Physics of the Earth of Russian Academy of
%Sciences, Bolshaya Gruzinskaya str. 10/1, Moscow, 123242, Russia
\author{Vasily Postnicov}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 123242, Russia}
\author{Alexey I. Samarin}
\altaffiliation{Computational Mathematics and Cybernetics, Lomonosov Moscow
  State University, Moscow, 119991, Russia}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 123242, Russia}
\author{Efim V. Lavrukhin}
\altaffiliation{Computational Mathematics and Cybernetics, Lomonosov Moscow
  State University, Moscow, 119991, Russia}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 123242, Russia}
\author{Marina V. Karsanina}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 123242, Russia}
\author{Mathieu Gravey}
\altaffiliation{University of Lausanne, Faculty of Geosciences and Environment,
  Institute of Earth Surface Dynamics, Lausanne, Switzerland}
\author{Kirill M. Gerke}
\email[E-mail:]{kg@ifz.ru}
\affiliation{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 123242, Russia}

\title{Evaluation of classical correlation functions from 2/3D images on CPU and
  GPU architectures: introducing CorrelationFunctions.jl}

@@

\section{Background and motivation}
Materials are ubiquitous both in nature and in industrial human activities, and
their physical properties are interconnected to
structure\cite{Torq_book}\cite{Sahimi_book}\cite{Adler_book}. The information
about the structure usually comes in the way of 2D and 3D images obtained by
microscopy, tomography and other related methodologies. This structural
information can be used for numerous types of analysis, including simulations of
different processes within studied materials or evaluation of their properties
using so-called pore-scale modelling. Unlike computational modelling techniques
a whole class of approximate methods exists including (rigorous) bounds that
allows very fast estimations of physical properties albeit the bounds themselves
are too wide to be applicable in majority of practical problems. Nonetheless, it
is noteworthy that such bounds are originally based on structural descriptors in
the form of correlation functions.

Two ways exist to obtain correlation functions for a given structure at hand:
measure them either experimentally with the help of scattering intensity or
from digital images. None is perfect, as the first approach suffers from the
limitation of CFs that can be obtained this way, while the second one provides
information with limited resolution or/and resolution to field-of-view ratio. In
addition to the resolution conundrum, the most useful imaging methods such as
X-ray computed tomography (XCT) and scanning electron microscopy (SEM) provide
gray-scale images (e.g., X-ray attenuation or electron back-scattering
distributions) due to their underlying physical principles and require labeling
of constituent phases (this is called segmentation) before computation of
CFs. If properly segmented, high-resolution digital images do provide a
possibility to compute any correlation function and, thus, possibility to
address numerous fundamental and practical research problems.

Correlation functions (CFs) as invaluable universal descriptors of structure are
utilized in a multitude of scientific disciplines: material sciences, rock
physics, soil physics and hydrology, cosmology and food engineering, biology and
many others. Computed from 2D and 3D images, CFs can be applied to:
\begin{enumerate}
  \item to characterize the morphology and representativeness via correlation
    lengths;
  \item to perform stochastic reconstructions from experimentally measured CFs
    or 2D to 3D reconstructions based on CFs that can be extracted from less
    dimensions or through penetrable sphere model;
  \item compare different structures to each other, including verification of
    stochastic reconstructions;
  \item compress structural information in the form of raw or parameterized CFs
    with the possibility to recover the structure using stochastic
    reconstruction
  \item describe structural dynamics under different boundary conditions;
  \item extract structural features for deep learning;
  \item fuse multi-scale images and structural information as obtained by
    different methods into a single digital model;
  \item quantify spatial heterogeneity.
\end{enumerate}

Stochastic reconstruction is a separate huge topic, as this approach allows to
solve an inverse problem and recover structure from known set of correlation
functions. This ability for recovery serves as the basis for majority of usages
in the list above. While it is necessary to compute CFs from digital images as a
target set for stochastic reconstructions, they rely more on efficient
recomputations or optimizations, e.g., re-evaluation during simulated
annealing. Such optimizations are not the part of the computational package and
are not considered in this work.

There are different types of correlation functions that in general describe some
probabilities. The main parameter of any correlation function is the number of
points that are utilized to evaluate such a probability. While so-called
$n$-point probability function\cite{Torq_book} will totally describe any
structure in the limit of $n \rightarrow$ number of voxels/pixels on the image,
computing or storing such a function is not practical. Based on information
content, it was shown that increasing the number of point $n > 2$ only
marginally improves the quality of the structure quantification and these
improvements decay with increasing $n$. For stochastic reconstruction purposes
computation of higher-order statistics will, arguably, not be balanced by
increased number of points, but this is the topic of active research. For
aforementioned reasons, we mainly focus on 2-point statistics, leaving the
possibility to include higher order CFs in the future work. The simplest 2-point
probability function ($S_2$ or autocorrelation) actually arise from small angle
scattering experiments and measures the probability that both ends of line
segment with a given length fall into the same phase. Other types include the
probability of a whole line segment to fall into the phase ($L_2$ function) or
to lie on the interface between phases ($F_{ss}$ function), and mainly originate
from aforementioned bounds on physical properties such as permeability and
elasticity. The general idea of computation of different 2-point CFs is shown in
\cref{fig:functions}. All major details and numerous analytical cases can be
found in seminal work of Torquato \cite{Torq_book}. To all CFs described in this
book and applicable to digital images we shall refer as to classical correlation
functions.

\begin{figure}[ht]
  \centering
  \includegraphics[width=.3\textwidth]{images/corr.png}
  \caption[]{{\small WTF?}}
  \label{fig:functions}
\end{figure}

While evaluation of correlation functions from 2D and 3D images was a topic of
numerous research papers, the information is highly fragmented, especially in
terms of algorithms and their implementation in the code. Some snippets of code
are available for some functions, but they lack general API and are usually
implemented in proprietary interpreted languages such as Matlab. Due to a recent
leap forward in GPU-based computations, some papers describe computations on
this architecture, but the code is usually not provided. All in all, the
efficient and open-source single API solution utilizing both CPU and GPU
architectures is currently absent.

The aim of this paper is to establish a general open-source solution allowing to
compute all classical correlation functions. This solution should be well
documented with all computational algorithms explained, it should also leverage
on recent advances in programming high-intensity computations and utilize both
CPU and GPU power. Our answer in the form of \code{CorrelationFunctions.jl}
package allows to compute numerous CFs easily under any operating system. The
manuscript describing the package is organized as follows: in \cref{scansec} we
provide all algorithmic details for CFs computation in directions (scanning
method). \cref{mapsec} presents all algorithmic details for computing full CFs
maps (map method, full 2-point statistics for a given image). In \cref{verisec}
we verify the all computations based on known analytical
solutions. \cref{effisec} deals with computational efficiency of our code, we
compare computational times for 2D and 3D images of different sizes for scanning
and map methods as evaluated using either CPU or GPU. In \cref{exasec} some
application examples are provided, including CFs evaluation for multi-phase
materials. Finally, \cref{sumsec} summarizes all results and outlines possible
future improvements.

@@

\section{Verifying our algorithms}
\label{verisec}
To verify our package we use random datasets of overlapping disks (the number of
dimensions $n = 2$) or balls ($n = 3$) with radii $R$ and centers placed in
points generated by Poisson point process with a parameter $\lambda$. For these
datasets correlation functions can be given in closed-form expressions for one-,
two- and three-dimensional cases. For the sake of testing we obtain a collection
of disks (resp. balls) in a square (resp. cube) with the side $s$ using the
following algorithm:
\begin{algorithmic}[1]
  \label{testdata}
  \Procedure{testdata}{$s, n, R, \lambda$}
    \State $nballs \gets poisson(\lambda s^n)$ \Comment{$poisson(\lambda)$
      generates random numbers with Poisson distribution with parameter
      $\lambda$.}
    \State Generate $nballs$ random points which are uniformly distributes in
    a cube ($n = 3$) or square ($n = 2$) with a side $s$
    \State Draw balls (disks) of radius $R$ in those points. Store the result in
    an array $output$.
    \State \textbf{return} $output$
  \EndProcedure
\end{algorithmic}

@@

\section{Efficiency of our algorithms}
\label{effisec}
To measure efficiecy of our algorithms we use a machine with the following
configuration:
\begin{itemize}
\item \textbf{CPU}: AMD Ryzen 5 1600X running at 2200~MHz
\item \textbf{Memory}: 32~Gb DDR4-2400 memory
\item \textbf{GPU}: Nvidia Tesla V100 with 32~Gb video memory
\end{itemize}

@@

\section{Examples and applications}
\label{exasec}
Text

\section{Summary}
\label{sumsec}
Text

@@

\begin{algorithmic}[1]
  \Procedure{countruns}{$array, len, phase$}
    \State $runslist \gets [\quad]$
    \State $runs \gets 0$
    \ForAll{$x \in array$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$array, runs, n$}
    \For{$idx = 1,n$}
      \State $array[idx] \gets array[idx] + runs$
      \State $runs \gets runs - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$array, first, last, n$}
    \State $sum \gets first + last$
    \State $n \gets \min(n, sum)$
    \For{$idx = 1,n$}
      \State $a \gets \min(idx - 1, first, last, sum - (idx - 1))$
      \State $array[idx] \gets array[idx] + a$
    \EndFor
  \EndProcedure
  \\
  \Procedure{l2}{$array, phase, maxlen, periodic$}
    \State $success \gets zeros(maxlen)$
    \State $total \gets zeros(maxlen)$
    \State $len \gets length(array)$
    \State $nupdates \gets \min(len, maxlen)$
    \State $runs \gets countruns(array, maxlen, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(success, run, \min(run, maxlen))$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(array)$
      \State $last \gets last(array)$
      \If{$first = last$}
        \State $updateperiodic(success, first, last, nupdates)$
      \EndIf
      \State Increment $total[\text{from} \, 1 \, \text{to} \, nupdates]$ by
      $len$
    \Else
      \State $updateruns(total, len, nupdates)$
    \EndIf
    \State \textbf{return} $success$ divided by $total$ element-wise.  
  \EndProcedure
\end{algorithmic}

\onecolumngrid
\bibliography{habib}
\bibliographystyle{plain}
\twocolumngrid

\end{document}
